---
title: "GLDS-366 GWAS Processed Associations"
output:
  html_notebook: default
  pdf_document: default
---

## Setup 

```{r}
# install.packages(c("systemfonts", "textshaping", "ragg"))
# install.packages("tidyverse")
```

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)
library(bnlearn)
library(Rgraphviz)
library(psych)
library(lavaan)
library(gnlearn)
library(mice)
library(GeneNet)
library(igraph)
```

```{r}
# Set data file paths
data_path <- paste0(getwd(), "/data/")

glds_366_gwas_path <- paste0(data_path, "GLDS-366_GWAS_processed_associations.csv")
```

```{r}
gwas <- read.csv(glds_366_gwas_path, header = T, stringsAsFactors = F)

numeric_cols <- gwas[sapply(gwas, is.numeric)]
non_numeric_cols <- gwas[sapply(gwas, function(x) !is.numeric(x))]
```

## Data Exploration
```{r}
str(gwas)
head(gwas)
summary(gwas)

missings <- sapply(gwas, function(x) sum(is.na(x)))
print(missings)

num_of_cols <- length(numeric_cols)

# Adjust grid based on number of numeric columns
par(mfrow = c(ceiling(sqrt(num_of_cols)), ceiling(sqrt(num_of_cols))))

# Histograms for numeric columns
for (col in names(numeric_cols)) {
  hist(numeric_cols[[col]], main = col, xlab = col, col = "lightblue", border = "black")
}

# Resetting graphical parameters for boxplots
par(mfrow = c(ceiling(sqrt(num_of_cols)), ceiling(sqrt(num_of_cols))))

# Boxplots for numeric columns
for (col in names(numeric_cols)) {
  boxplot(numeric_cols[[col]], main = col, col = "lightblue", border = "black")
}

if (length(numeric_cols) > 0) {
  cat("The following columns are numeric:\n")
  print(names(numeric_cols))
}

# Non-numeric columns
if (length(non_numeric_cols) > 0) {
  cat("The following columns are non-numeric:\n")
  print(names(non_numeric_cols))
}
```

```{r}
range_values <- range(non_numeric_cols$"Bgd_X.ray_8")
range_values # NA NA

range_values <- range(non_numeric_cols$"FociPerGy_X.ray_8")
range_values # NA NA

range_values <- range(non_numeric_cols$"chromosome")
range_values # 1 Y

unique_values <- unique(non_numeric_cols$"chromosome")
unique_values # "1"  "2"  "3"  "4"  "5"  "6"  "7"  "8"  "9"  "10" "11" "12" "13" "14" "15" "16" "17" "18" "19" "M"  "X"  "Y"
```

```{r}
head(gwas)

numeric_cols_clean_comb_all <- numeric_cols_clean_comb
numeric_cols_clean_comb_all$position.b38 <- gwas$position.b38

numeric_cols_clean_comb_all$chromosomePosition <- paste(numeric_cols_clean_comb_all$chromosome, "-", numeric_cols_clean_comb_all$position.b38, sep = "")

head(numeric_cols_clean_comb_all$chromosomePosition)
sum(table(numeric_cols_clean_comb_all$chromosomePosition) == 1)
dim(numeric_cols_clean_comb_all)
x <- table(numeric_cols_clean_comb_all$chromosomePosition)

x[x > 1]
```

## Data Preprocessing
```{r}
char_to_numer <- function(x) {
  if (x == "M") {
    return(20)
  } else if (x == "X") {
    return(21)
  } else if (x == "Y") {
    return(22)
  } else {
    return(as.numeric(x))
  }
}

numeric_cols$chromosome <- sapply(non_numeric_cols$"chromosome", char_to_numer)

range_values <- range(numeric_cols$chromosome)
range_values # 1 22

head(numeric_cols)
```

### All Numerical
```{r}
# Delete any rows with NA values
numeric_cols_clean <- drop_na(numeric_cols)
head(numeric_cols_clean)
dim(numeric_cols) # 134593     61
dim(numeric_cols_clean) # 56191    62
```

### Subset by n_#
```{r}
# Subset by n_#
suffixes <- c("n_4", "n_8", "n_24", "n_48")
df_list <- list()

# Loop through each suffix to create the dataframes
for (suffix in suffixes) {
  df_list[[suffix]] <- numeric_cols_clean %>%
    select(ends_with(c(suffix, paste0(suffix, ".1"))))
}

# Access each dataframe by using the suffixes
numeric_cols_clean_n4 <- df_list[["n_4"]]
numeric_cols_clean_n8 <- df_list[["n_8"]]
numeric_cols_clean_n24 <- df_list[["n_24"]]
numeric_cols_clean_n48 <- df_list[["n_48"]]

head(numeric_cols_clean_comb_n4)
```

### Subset by n_# & combined with and without .1
```{r}
# Average columns with the same name but with or without the .1 suffix
# Extract column names that end with .1
cols_with_dot <- grep("\\.1$", names(numeric_cols_clean), value = TRUE)

# Extract the corresponding original column names (without .1)
original_cols <- sub("\\.1$", "", cols_with_dot)

numeric_cols_clean_comb <- numeric_cols_clean

# For each base column name, average with its .1 counterpart
for (col in original_cols) {
  if (col %in% names(numeric_cols_clean_comb) && paste0(col, ".1") %in% names(numeric_cols_clean_comb)) {
    numeric_cols_clean_comb[[col]] <- rowMeans(numeric_cols_clean_comb[, c(col, paste0(col, ".1"))], na.rm = TRUE)
  }
}

# Drop columns with .1 suffix
numeric_cols_clean_comb <- numeric_cols_clean %>% select(-ends_with(".1"))

head(numeric_cols_clean_comb)

# Subset by n_#
suffixes <- c("n_4", "n_8", "n_24", "n_48")
df_list <- list()

for (suffix in suffixes) {
  df_list[[suffix]] <- numeric_cols_clean_comb %>%
    select(ends_with(c(suffix, paste0(suffix, ".1"))))
}

numeric_cols_clean_comb_n4 <- df_list[["n_4"]]
numeric_cols_clean_comb_n8 <- df_list[["n_8"]]
numeric_cols_clean_comb_n24 <- df_list[["n_24"]]
numeric_cols_clean_comb_n48 <- df_list[["n_48"]]

head(numeric_cols_clean_comb_n4)
```

### Subset by prefix of Bgd or FociPerGy & combined with and without .1
```{r}
# Combine dfs with averaged columns
complete_numeric_clean_comb_df <- bind_cols(df_list)
head(complete_numeric_clean_comb_df)

# Subset by prefix of Bgd or FociPerGy
prefixes <- c("Bgd", "FociPerGy")
df_list <- list()

# Loop through each prefix to create the dataframes
for (prefix in prefixes) {
  df_list[[prefix]] <- complete_numeric_clean_comb_df %>%
    select(starts_with(prefix))
  head(df_list[[prefix]])
}

# Access each dataframe by using the prefixes
complete_numeric_clean_comb_bgd <- df_list[["Bgd"]]
head(complete_numeric_clean_comb_bgd)

complete_numeric_clean_comb_foci <- df_list[["FociPerGy"]]
head(complete_numeric_clean_comb_foci)
```

### Subset by prefix of Bgd or FociPerGy & combined with and without .1 plus chromosome
```{r}
# Combine dfs with averaged columns
complete_numeric_clean_comb_bgd_chromosome <- complete_numeric_clean_comb_bgd
complete_numeric_clean_comb_bgd_chromosome$chromosome <- numeric_cols_clean$chromosome

head(complete_numeric_clean_comb_bgd_chromosome)

complete_numeric_clean_comb_foci_chromosome <- complete_numeric_clean_comb_foci
complete_numeric_clean_comb_foci_chromosome$chromosome <- numeric_cols_clean$chromosome

head(complete_numeric_clean_comb_foci_chromosome)
```

## NoDag

### Numerical Columns - All
```{r}
head(cor(numeric_cols_clean))

gwas_numer_dag <- nodag(numeric_cols_clean)
plot(gwas_numer_dag)
```

### Numerical Columns - Subset by n_#
```{r}
gwas_numer_n4_dag <- nodag(numeric_cols_clean_n4)
plot(gwas_numer_n4_dag)

gwas_numer_n8_dag <- nodag(numeric_cols_clean_n8)
plot(gwas_numer_n8_dag)

gwas_numer_n24_dag <- nodag(numeric_cols_clean_n24)
plot(gwas_numer_n24_dag)

gwas_numer_n48_dag <- nodag(numeric_cols_clean_n48)
plot(gwas_numer_n48_dag)
```

### Numerical Columns - Subset by n_# & combined with and without .1
```{r}
gwas_numer_n4_comb_dag <- nodag(numeric_cols_clean_comb_n4)
plot(gwas_numer_n4_comb_dag)

gwas_numer_n8_comb_dag <- nodag(numeric_cols_clean_comb_n8)
plot(gwas_numer_n8_comb_dag)

gwas_numer_n24_comb_dag <- nodag(numeric_cols_clean_comb_n24)
plot(gwas_numer_n24_comb_dag)

gwas_numer_n48_comb_dag <- nodag(numeric_cols_clean_comb_n48)
plot(gwas_numer_n48_comb_dag)
```

### Numerical Columns - Subset by prefix of Bgd or FociPerGy & combined with and without .1
```{r}
gwas_all_numer_comb_bgd_dag <- nodag(complete_numeric_clean_comb_bgd)
plot(gwas_all_numer_comb_bgd_dag)

gwas_all_numer_comb_foci_dag <- nodag(complete_numeric_clean_comb_foci)
plot(gwas_all_numer_comb_foci_dag)
```

### Subset by prefix of Bgd or FociPerGy & combined with and without .1 plus chromosome
```{r}
complete_numeric_clean_comb_bgd_chromosome_dag <- nodag(complete_numeric_clean_comb_bgd_chromosome)
plot(complete_numeric_clean_comb_bgd_chromosome_dag)

complete_numeric_clean_comb_foci_chromosome_dag <- nodag(complete_numeric_clean_comb_foci_chromosome)
plot(complete_numeric_clean_comb_foci_chromosome_dag)
```